---
title: "D3. Project development: descriptive analysis"
output: word_document
editor_options: 
  chunk_output_type: console
---
#7TotaldescriptivaClean5.Rmd
#install.packages("rmarkdown")
#library("rmarkdown")

## Introduction
This document provides the initial univariate descriptive statistics of the raw variables in the hotel booking dataset.

#loading libraries
install.packages("ggplot2")
library(ggplot2) 
install.packages("dplyr")
library(dplyr)

#WARNING: data must have been properly declared before (factors, dates...)

#Set the folder where the word file is going to be generated
```{r, echo=FALSE} 

setwd("/home/alex/Pictures")
dd<- read.table("hotel_booking_5000_rows.csv",header=T, sep=",", dec=".")

```  

#without including the R instruction in the final document
```{r, echo=FALSE}              
class(dd)
```

#Get dimensions of the dataset
````{r, echo=FALSE} 
dim(dd)
n<-dim(dd)[1]
K<-dim(dd)[2]

n
K
````

#Check the variables 
```{r, echo=FALSE}  
names(dd)
str(dd)
head(dd) 
summary(dd)
```

#Decide if you need to declare some more factor or date
```{r, echo=FALSE}  
#para la conversión numérica de los meses 
library(dplyr)

month_mapping <- setNames(1:12, month.name)
dd <- dd %>%
  mutate(
    arrival_date_year = as.numeric(arrival_date_year),
    arrival_date_month = as.numeric(month_mapping[arrival_date_month]),
    arrival_date_day_of_month = as.numeric(arrival_date_day_of_month)
  )


# creación arrival_date, combinación 3 columnas
dd <- dd %>%
  mutate(
    arrival_date = as.Date(
      paste(arrival_date_year, arrival_date_month, arrival_date_day_of_month, sep = "-"),
      format = "%Y-%m-%d"
    )
  )

dd <- dd %>%
  mutate(
    hotel = as.factor(hotel),
    is_canceled = as.factor(is_canceled),
    meal = as.factor(meal),
    country = as.factor(country),
    market_segment = as.factor(market_segment),
    distribution_channel = as.factor(distribution_channel),
    is_repeated_guest = as.factor(is_repeated_guest),
    reserved_room_type = as.factor(reserved_room_type),
    assigned_room_type = as.factor(assigned_room_type),
    deposit_type = as.factor(deposit_type),
    customer_type = as.factor(customer_type),
    reservation_status = as.factor(reservation_status),
    company = as.factor(company),
    arrival_date_week_number = as.factor(arrival_date_week_number)
  )

dd <- dd %>%
  select(-c(name, email, phone.number, credit_card, arrival_date_year, arrival_date_month, arrival_date_day_of_month))

K<-dim(dd)[2]

str(dd)

```

#Calculation of mean, median and standard deviation
```{r, echo=FALSE} 
numeric_cols <- dd %>%
  select(where(is.numeric))

summary_numcols <- numeric_cols %>%
    summarise_all(list(
      mean = ~mean(., na.rm = TRUE),
      median = ~median(., na.rm=TRUE),
      sd = ~sd(., na.rm = TRUE),
      var = ~var(., na.rm = TRUE)
    ))

print(summary_numcols)

```

#Given scheme for descriptive analyis
```{r, echo=FALSE} 

descriptiva<-function(X, nom){
  if (!(is.numeric(X) || class(X)=="Date")){ 
    frecs<-table(as.factor(X), useNA="ifany")
    proportions<-frecs/n
    #ojo, decidir si calcular porcentages con o sin missing values
    pie(frecs, cex=0.6, main=paste("Pie of", nom))
    barplot(frecs, las=3, cex.names=0.7, main=paste("Barplot of", nom), col=listOfColors)
    print(paste("Number of modalities: ", length(frecs)))
    print("Frequency table")
    print(frecs)
    print("Relative frequency table (proportions)")
    print(proportions)
    print("Frequency table sorted")
    print(sort(frecs, decreasing=TRUE))
    print("Relative frequency table (proportions) sorted")
    print(sort(proportions, decreasing=TRUE))
   }else{
     if(class(X)=="Date"){
       print(summary(X))
       print(sd(X))
       #decide breaks: weeks, months, quarters...
       hist(X,breaks="weeks")
     }else{
       hist(X, main=paste("Histogram of", nom))
       boxplot(X, horizontal=TRUE, main=paste("Boxplot of",nom))
       print("Extended Summary Statistics")
       print(summary(X))
       print(paste("sd: ", sd(X, na.rm=TRUE)))
       print(paste("vc: ", sd(X, na.rm=TRUE)/mean(X, na.rm=TRUE)))
      }
   }
}

dataset<-dd
actives<-c(1:K)

```

#Basic descriptive analysis for numerical variables
#(decide the maximum number of colors you can need in a graph based on your metadata file)
```{r, echo=FALSE} 
listOfColors<-rainbow(39)

par(ask=TRUE)

for(k in actives){
  print(paste("variable ", k, ":", names(dd)[k] ))
  descriptiva(dd[,k], names(dd)[k])
}
par(ask=FALSE)
```

#Bivariate analysis
```{r, echo=FALSE} 

temp_numeric_cols <- numeric_cols

temp_numeric_cols <- temp_numeric_cols %>%
  select(-c(previous_cancellations, previous_bookings_not_canceled, booking_changes, agent, required_car_parking_spaces, total_of_special_requests)
         )


correlation_matrix <- cor(temp_numeric_cols, use = "complete.obs")  # Handles missing values
print(correlation_matrix)
correlation_matrix |> 
  round(2) |> 
  write.csv("correlation_matrix.csv", row.names = TRUE)

par(ask=TRUE)
library(ggplot2)
ggplot(dd, aes(x = stays_in_weekend_nights, y = stays_in_week_nights)) +
  geom_point(alpha = 0.5, color = "steelblue", na.rm = TRUE) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Relationship between Weekend and Weekday Stays",
       x = "Stays in Weekend Nights",
       y = "Stays in Week Nights") +
  theme_minimal()

# Scatter plot for lead_time vs arrival_date_week_number
ggplot(dd, aes(x = as.numeric(as.character(arrival_date_week_number)), y = lead_time)) +
  geom_point(alpha = 0.4, color = "darkgreen", na.rm = TRUE, position = position_jitter(width = 0.2)) +
  geom_smooth(method = "loess", color = "orange", se = FALSE) +
  labs(title = "Lead Time vs Booking Week",
       x = "Arrival Week Number (1-53)",
       y = "Lead Time (Days)") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1, 53, 4))

par(ask=FALSE)

```

#per exportar figures d'R per programa
#dev.off()
#png(file=mypath,width = 950, height = 800, units = "px")
#dev.off()



